The role of $4f$ electrons in the stopping power of hafnium

### 2020-02-21 hafnio_ale.tex
intro and experiment revised

%=============================================================================================================================================================================================
%=============================================================================================================================================================================================
\begin{abstract}
 The stopping power of protons through Hf foil has been studied both experimentally and theoretically. The measurements were performed at the Laboratory of Accelerators and X-Ray Diffraction in Lisbon, by using the transmission method on self-supporting stopping material, for (0.6-2.5) MeV protons. The theoretical developments involved fully relativistic atomic structure calculations for %this Z=72 target 
 Hf, which required the solution of the Dirac equation. The shell-wise local plasma approximation (SLPA) was employed to describe the energy transferred to the bound 1s-4f electrons, and the outer four electrons were considered as a free electron gas (FEG). %We contemplated two different models for the FEG: the screened potential with cusp condition (SPCC) for energies below that of the plasmon excitation, and the dielectric formalism, for energies around the stopping maximum and above. 
 We found the relativistic description of the 4f-shell and the screening between $4f$ and $5p$ electrons are decisive around the stopping maximum. %for energies above 40 keV.
 Present theoretical and experimental results are in very good agreement in the energy region of the new measurements. However, our theoretical stopping cross sections show %interesting 
 substantial differences with the semi-empirical values by SRIM2013 and ICRU-49 at intermediate to low energies. Our calculations suggest the stopping maximum %around 40 eV$/(10^{15}$ atom/cm$^2)$ at 70 keV, which is
 higher and shifted to lower energies than previous predictions. Future measurements below 100 keV would be decisive for a better knowledge of the stopping maximum and below.
\end{abstract}

@@

For impact energies above a few keV/amu, mono-energetic charged particles penetrating a foil of any material lose their energy through a series of consecutive inelastic collisions, mainly with target electrons \cite{Chu01,Sigmund}. The information given by the energy loss process is essential not only to have a better knowledge of the physics behind the fundamental interactions, but also because it plays an important role in many applied fields such as materials science, nuclear physics, ionic implantation and radiotherapy \cite{Sigmund,Schardt}. Moreover, experimental values on ion mean energy loss per unit path $S(E)$ are of crucial relevance in order to check the reliability of semi-empirical models and to determine some key parameters 
%included in theoretical approaches of stopping power %\cite{Diwan,Damache01,Damache02,Damache03,Damache04}. 
\cite{Diwan,Damache04,Damache02}. However, very often, the experimental data available is rather scarce which is particularly troublesome when the material under study corresponds to an element of low occurrence on the Earth's upper crust, such as Hf.

So far, only one experimental work has been published regarding the stopping power cross section of pure hafnium for protons \cite{Sirotinin}, while more attention has been recently given to studies involving hafnium oxide due to its practical use \cite{Abril,Behar,Primetzhofer}. It is well known that significant attention has been paid in recent years to transition metaloxides such as HfO$_2$ for their potential as alternative gate dielectrics to replace SiO$_2$ for the future generation of nano-electronics with less than 45 nm gate length \cite{Choi,Robertson}. In this regard, some important physical properties of the above mentioned metal-oxide films depend on its thickness, which is often measured by using Rutherford Backscattering Spectrometry \cite{Alfassi01, Tesmer01}, a methodology that relies heavily on the determination of the stopping power of ion beams in the material of interest.

In this study we report stopping power cross section experimental values over the incident energy range (0.6-2.5) MeV for protons crossing self-supported Hf thin-film by using the transmission method. Our aim is not only to upgrade stopping power data compilations \cite{HPaul03,mondim17} but also to provide useful information about the processes governing the slowing down of protons in multielectronic targets. In the rare earth metals, the 4f electrons play a relevant role in the stopping power, being the first shell of bound electrons below the conduction band. As already noted \cite{Roth17}, the FEG shows unexpected %a peculiar
behaviour in these elements that states doubts on the proper description. In the case of Hf we found the contribution of the 4f-shell decisive even at impact energies around the stopping maximum, as we will show later in this work.

The theoretical approach used in this work requires 
the relativistic wave functions and binding energies of Hf, and considering 4 electrons per atom in the free electron gas (FEG) \cite{mendez2019}. The shell-wise local plasma approximation (SLPA) \cite{mon13} was employed to describe the energy transferred to the bound 1s-4f electrons, and two different models for the FEG: the screened potential with cusp condition model (SPCC) \cite{mon17}, which is non-linear binary formalism%,  for energies below that of the plasmon excitation
; and the Mermin-Lindhard dielectric formalism (ML) \cite{Mermin}, for energies around the stopping maximum and above.  On the other hand, Hf has the extra interest of the filled $4f$-subshell (with 14 electrons) as the main contribution below the FEG, causing the stopping cross sections to be very sensitive to a good description of this shell. The screening among the $4f$ and $5p$ electrons has been considered and found to play a major role within the SLPA calculations.

The experimental details and data are given in section \ref{experiment}, while the theoretical method is explained in section \ref{theory}. Present theoretical and experimental values are finally compared to the available experimental values \cite{Sirotinin},  the ICRU-49 calculations \cite{ICRU49} and the SRIM-2013 package \cite{Ziegler01}. Conclusions and discusions are given in section \ref{conclusion}.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{Fig01.eps}
\caption{RBS spectrum for $E_{avg} = 921,1$ keV protons on hafnium sample which is subsequently used to determine the energy loss in the foil.}
\label{F01}
\end{figure}

@@

\section{Experimental arrangements}
\label{experiment}
\subsection{Accelerator and scattering Chamber}
The procedure used in this work to obtain stopping power data is essentially the same as described in reference \cite{Miranda01}. However, this time the measurements were made at the IST/LATR
(Laboratory of Accelerators and X-Ray Diffraction) in Lisbon. This facility uses a 2.5 MV Van de Graaff accelerator to deliver $^1$H$^+$ primary ion beams through a series of electrostatic lenses and collimators onto a thin Au/SiO$_2$ sample which is used as a scattering center. This sample was placed in the center of a RBS/C scattering chamber where a high vacuum (pressure of
$\sim$10$^{-6}$ Torr) was maintained during the measurements. The beam current on sample was kept at around 5.0 nA in order to attain sufficient statistics in each particle spectrum.
By using a beam spot of about 1.0 mm in diameter, a solid angle of 11.4 msr was attained. The overall energy resolution (FWHM) of the detection system was about 15 keV relative to 5.486 MeV alpha particles from a $^{241}$Am source.

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Target}
The stopping material under analysis was a hafnium foil with nominal thickness of 1.0 $\mu$m and 99.95\% purity which was supplied by Lebow Company \cite{Lebow}. However, a more precise thickness value was achieved by measuring the energy loss of alpha particles coming from a calibrated ($^{239}$Pu, $^{241}$Am, $^{244}$Cm) source. From the alpha spectra with and without the Hf foil interposed, the characteristic energy shift $\delta$E was measured and then combined with the stopping power for 5.486 MeV alphas on hafnium (55.69 eV/$10^{15}$ at/cm$^2$) found in Ref.\cite{Ziegler01} in order to obtain an areal density of (4,13$\pm$0.21)$\times 10^{19}$ at/cm$^2$ which corresponds to a thickness of 0.920$\pm$0.046 $\mu$m.
%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Energy loss measurement}
Once the beam impinges on the Au/SiO$_2$ sample, protons are backscattered towards a Si surface barrier detector located at 140$^{\circ}$ relative to the initial beam direction. Fig.\ref{F01} shows two particle spectra where ion energies $E_2$ and $E_1$ are associated to a placed and removed hafnium sample, respectively. Both energy distributions were fitted by gaussian functions to obtain the mean energy and width (FWHM) of the peaks \cite{Sun01}, and from the difference between these two peak positions in the spectrum, the total energy loss $\Delta E = (E_1 - E_2)$ in the foil was calculated. As established in previous studies \cite{Miranda01, Damache02}, the experimental stopping powers cross sections $\varepsilon (E) $ are determined at some mean energy $E_{avg}$ by measuring the ion energy losses $\Delta E$ within the investigated Hf foil which has a mean thickness denoted by $\Delta x$. In this way, only when the energy loss fraction $\Delta E/E_{avg}$ across the Hf foil is not exceeding 20\%
it is possible to define the stopping cross section by \cite{Raisanen01,Schulz01}:
\begin{equation}\label{eq:stcross}
 \varepsilon (E) = \frac{S(E)}{N} = -\frac{dE}{N dx} \approx -\frac{\Delta E}{N \Delta x},
\end{equation}
where $N$ denotes the atomic number density (atoms cm$^{-3}$) of the material under study. When this condition was not fulfilled, a small correction to the mean energies $E_{avg}$ was applied in order to account for the nonlinear dependence on ion energy of stopping powers \cite{Chilton,Rajatora}.

\section{Theoretical Method} \label{theory}
The energy loss of ions in metal targets responds to different physical mechanisms, depending on the impact ion velocity. At low velocities the binary collisions are responsible of the loss of energy by the ion. The main contribution is the ionization of electrons of the metal conduction band, which is well approximated by a free electron gas (FEG) of Fermi velocity $v_F$. Instead, above certain velocity (i.e. $v\geq 1.5 \ v_F$) , not only binary but also collective excitations (plasmons) are possible \cite{mon17}. At high energies, not only the FEG contributes to the stopping power but also the bound electrons. The method used in this work combines a description for the interaction with the valence (or conduction) electrons as a FEG, and a different one for the interaction with the bound electrons.

We used the SPCC model \cite{mon17} to describe the stopping power of low velocity charged particles in the FEG. It is a non-perturbative binary collisional model, thus valid at energies below that of plasmon excitations.  The SPCC \cite{mon17} is based on a screened central potential with cusp condition of the electronic density close to the projectile. It proved to give a good description of the induced electron density even for negative projectiles \cite{mon17}, and reproduces the low velocity proton-antiproton differences in the stopping power (Barkas effect). The SPCC formalism only depends on the Wigner-Seitz radio, $r_S$, which is a measure of the electronic density of the FEG. For metals of well-known value of $r_S$, the SPCC describes very well the low energy experimental stopping data \cite{mon17}, agreeing with the DFT results by Echenique and coworkers \cite{eche81,nagy89} at $v=0$.

%In the case of 
Hafnium (Z=72, [Xe] $4f^{14} \ 6s^2 \ 5d^2$), belongs to the first groups of transition metals, with  4 electrons as FEG (theoretical $r_S=2.14$ a.u.) and $1s-4f$ electrons bound. We checked the $r_S$ value with the experimental one obtained from measured energy loss function by Lynch \textit{et al} \cite{lynch75}. The experimental plasmon energy of Hf is $\hbar\omega_P \approx 15.8$ eV, with a width at half maximum $\delta \approx 4.4$ eV, and $r_S \approx 2.07$ a.u. \cite{lynch75}. Following \cite{mon17}, the difference of less than $5\%$ between theoretical and experimental $r_S$ assess Hf as a canonical target.

Above certain impact velocity the plasmon contribution is important (i.e. around and above the stopping maximum). An interesting value for our analysis is the minimum impact velocity to excite
plasmons, $v_P$. In the dielectric formalism this value can be obtained as $v_P\ \approx \ v_F[1+(3 \pi\ v_F)^{-1/2} ]$ \cite{suppression}. To describe the energy loss considering collective and binary excitation we resort to the ML dielectric formalism \cite{Mermin}, which is a linear response, perturbative approximation, so it depends on the square of the ion charge. In this formalism, the response of target electrons to the ion passage is described through the quantum dielectric function, depending on the characteristic $r_S$ and $\delta$ parameters of the FEG.

@@

For the stopping power due to bound electrons, the SLPA \cite{mon17,mon13} is employed. It is worth to mention that the only inputs for the SLPA are the space-dependent densities of each shell in the ground state, and their binding energies. Collective processes and screening among electrons are included. Hafnium is a relativistic target, therefore the wave functions and binding energies must be obtained as solutions of Dirac equation  \cite{mendez2019}.

Figure \ref{Binding_E} (a) shows non relativistic and our relativistic binding energies $E_{nl\pm}$, where $\pm=j\pm1/2$.
We compare our results with experimental measurements on solid Hf \cite{williams1995} from $1s$ to $4f_{\pm}$. We notice that not only the deep shells require relativistic calculations, but also the outer $5p$ and $4f$ shells. Figure \ref{Binding_E} (b) displays the relative error of non relativistic and relativistic energies with respect to the experimental data in logarithmic scale. This figure shows very clearly the disability of non-relativistic calculations to describe the experimental data, which surprisingly worsens from the inner to the outer shells. 
%It is clear in this figure the disability of non-relativistic binding energies to describe the experimental data, which surprisingly worsens from the inner to the outer-shells.
%We used here the theoretical binding energies and wave functions of Hf calculated in \cite{mendez2019}. 
%The present theoretical description of Hf considers a FEG of $r_S=2.14$ and 1s-4f electrons as bound.

>>>>

%=============================================================================================================================================================================================
%=============================================================================================================================================================================================
\begin{abstract}
The stopping power of protons through Hf foil has been studied both experimentally and theoretically. 
The measurements were performed at the Laboratory of Accelerators and X-Ray Diffraction in Lisbon, by using the transmission method on self-supporting stopping material, for (0.6-2.5) MeV protons.
The theoretical developments involved fully relativistic atomic structure calculations for Hf, which required the solution of the Dirac equation. 
The shell-wise local plasma approximation (SLPA) was used to describe the energy transferred to the bound 1s-4f electrons, and the outer four electrons were considered as a free electron gas (FEG). 
 We found the relativistic description of the 4f-shell, and the screening between $4f$ and $5p$ electrons are decisive around the stopping maximum. 
The present theoretical and experimental results have very good agreement in the energy region of the new measurements. 
However, our theoretical stopping cross sections show substantial differences with the semi-empirical models, such as SRIM2013 and ICRU-49, at intermediate to low energies. 
Our calculations suggest the stopping maximum to be larger, and shifted to lower energies, respect to these semi-empirical models. Future measurements below 100 keV would be required to validate our predictions.
\end{abstract}

@@

For impact energies above a few keV/amu, mono-energetic charged particles penetrating a foil of any material lose their energy through a series of consecutive inelastic collisions, mainly with target electrons \cite{Chu01,Sigmund}. 
The information given by the energy loss process is essential not only to have a better knowledge of the phenomena behind the fundamental interactions but also because it plays an important role in many applied fields such as materials science, nuclear physics, ionic implantation, and radiotherapy \cite{Sigmund,Schardt}. 
Experimental data in ion mean energy loss per unit path $S(E)$ is of crucial relevance to check the reliability of semi-empirical models and to determine some key parameters \cite{Diwan,Damache04,Damache02}. 
However, often, the experimental data available is rather scarce, which is troublesome when the material under study corresponds to an element of low occurrence on the Earth's upper crust, such as Hf.

So far, only one experimental work has been published regarding the stopping power cross section of pure hafnium for protons \cite{Sirotinin}, while more attention has been recently given to studies involving hafnium oxide because of its practical use \cite{Abril,Behar,Primetzhofer}. 
It is well known that significant attention has been paid in recent years to transition metal-oxides such as HfO$_2$ because of their potential as alternative gate dielectrics to replace SiO$_2$ for the future generation of nano-electronics with less than 45 nm gate length \cite{Choi,Robertson}. 
Some important physical properties of the above mentioned metal-oxide films depend on its thickness, which is often measured by using Rutherford Backscattering Spectrometry \cite{Alfassi01, Tesmer01}, a method that relies heavily on the determination of the stopping power of ion beams in the material of interest.

In this study, we report experimental stopping power cross section values over the incident energy range (0.6-2.5) MeV for protons crossing self-supported Hf thin-film by using the transmission method.
We aim not only to upgrade stopping power data compilations \cite{HPaul03,mondim17} but also to provide useful information about the processes governing the slowing down of protons in multi-electronic targets. 
In the rare earth metals, the $4f$ electrons play a relevant role in the stopping power, being the first shell of bound electrons below the conduction band. As already noted \cite{Roth17}, the free electron gas (FEG) shows unexpected behavior in these elements, which casts doubts on its proper description. With Hf, we found the contribution of the 4f-shell decisive even at impact energies around the stopping maximum, as shown later.

The theoretical approach implemented in this work uses the shell-wise local plasma approximation (SLPA) \cite{mon13} to describe the energy transferred to the bounds 1s-4f electrons, and two different models for the FEG: the screened potential with cusp condition model (SPCC) \cite{mon17}, which is non-linear binary formalism; and the Mermin-Lindhard dielectric formalism (ML) \cite{Mermin}, for energies around the stopping maximum and above. 
Our model requires the relativistic description of the wave functions and binding energies of Hf \cite{mendez2019}, where we consider 4 electrons per atom in the FEG. 
The 14 electrons occupying the 4f subshell of Hf makes up the main contribution to the stopping below the FEG, which causes the cross sections to be very sensitive to a good description of this shell. 
The screening between the 4f and 5p electrons has been considered, and it has been found to play a major role within the SLPA predictions.

%The theoretical approach used in this work requires the relativistic wave functions and binding energies of Hf, and considering 4 electrons per atom in the free electron gas (FEG) \cite{mendez2019}. The shell-wise local plasma approximation (SLPA) \cite{mon13} was employed to describe the energy transferred to the bound 1s-4f electrons, and two different models for the FEG: the screened potential with cusp condition model (SPCC) \cite{mon17}, which is non-linear binary formalism; and the Mermin-Lindhard dielectric formalism (ML) \cite{Mermin}, for energies around the stopping maximum and above.  On the other hand, Hf has the extra interest of the filled $4f$-subshell (with 14 electrons) as the main contribution below the FEG, causing the stopping cross sections to be very sensitive to a good description of this shell. The screening among the $4f$ and $5p$ electrons has been considered and found to play a major role within the SLPA calculations.

The experimental details and data are given in Section \ref{experiment}, while the theoretical method is explained in Section \ref{theory}. Finally, our theoretical approach and experimental measurements are compared to the available experimental values \cite{Sirotinin}, the ICRU-49 calculations \cite{ICRU49}, and the SRIM-2013 package \cite{Ziegler01}. Conclusions and discussions are given in Section \ref{conclusion}.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{Fig01.eps}
\caption{RBS spectrum for $E_{\mathrm{avg}} = 921.1$ keV protons on hafnium sample, which is subsequently used to determine the energy loss in the foil.}
\label{F01}
\end{figure}

@@

\section{Experimental arrangements}
\label{experiment}
\subsection{Accelerator and scattering Chamber}
The procedure carried out to obtain the stopping power data in this work is essentially the same as the one detailed in Ref. \cite{Miranda01}. However, the present measurements were performed at the IST/LATR (Laboratory of Accelerators and X-Ray Diffraction) in Lisbon. This facility uses a 2.5 MV Van de Graaff accelerator to deliver $^1$H$^+$ primary ion beams through a series of electrostatic lenses and collimators onto a thin Au/SiO$_2$ sample which is used as a scattering center. This sample was placed in the center of an RBS/C scattering chamber, where a high vacuum (pressure of $\sim$10$^{-6}$ Torr) was maintained during the measurements. The beam current on the sample was kept at around 5.0 nA to attain sufficient statistics in each particle spectrum. 
By using a beam spot of about 1.0 mm in diameter, a solid angle of 11.4 msr was attained. The overall energy resolution (FWHM) of the detection system was about 15 keV relative to 5.486 MeV alpha particles from a $^{241}$Am source.

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Target}
The stopping material under analysis was a hafnium foil with a nominal thickness of 1.0 $\mu$m and 99.95\% purity, which was supplied by Lebow Company \cite{Lebow}. However, a more precise thickness value was achieved by measuring the energy loss of alpha particles coming from a calibrated ($^{239}$Pu, $^{241}$Am, $^{244}$Cm) source. From the alpha spectra with and without the Hf foil interposed, the characteristic energy shift $\delta$E was measured and then combined with the stopping power for 5.486 MeV alphas on hafnium (55.69 eV/$10^{15}$ at/cm$^2$) found in Ref.\cite{Ziegler01} to obtain an areal density of ($4.13\pm0.21$)$\times 10^{19}$ at/cm$^2$, which corresponds to a thickness of $0.920\pm0.046 \mu$m.

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Energy loss measurement}
Once the beam impinges on the Au/SiO$_2$ sample, protons are backscattered towards a Si surface barrier detector located at 140$^{\circ}$ relative to the initial beam direction. Fig.\ref{F01} shows two-particle spectra, where ion energies $E_2$ and $E_1$ are associated to a placed and removed hafnium sample, respectively. Both energy distributions were fitted by Gaussian functions to obtain the mean energy and width (FWHM) of the peaks \cite{Sun01}, and from the difference between these two peak positions in the spectrum, the total energy loss $\Delta E = (E_1 - E_2)$ in the foil was calculated. As established in previous studies \cite{Miranda01, Damache02}, the experimental stopping power cross sections $\varepsilon (E) $ are determined at some mean energy $E_{\mathrm{avg}}$ by measuring the ion energy losses $\Delta E$ within the investigated Hf foil, which has a mean thickness denoted by $\Delta x$. In this way, only when the energy loss fraction $\Delta E/E_{\mathrm{avg}}$ across the Hf foil is not exceeding 20\% it is possible to define the stopping cross section by \cite{Raisanen01,Schulz01}
\begin{equation}\label{eq:stcross}
 \varepsilon (E) = \frac{S(E)}{N} = -\frac{dE}{N dx} \approx -\frac{\Delta E}{N \Delta x}\,,
\end{equation}
where $N$ denotes the atomic number density (atoms cm$^{-3}$) of the material under study. When this condition was not fulfilled, a small correction to the mean energies $E_{\mathrm{avg}}$ was applied in order to account for the nonlinear dependence on ion energy of stopping powers \cite{Chilton,Rajatora}.

\section{Theoretical Method} 
\label{theory}
The energy loss of ions in metal targets responds to different physical mechanisms, depending on the impact ion velocity. At low velocities, the binary collisions are responsible for the loss of energy by the ion. The main contribution is the ionization of electrons in the metal conduction band, which is well approximated by a free electron gas (FEG) of Fermi velocity $v_F$. Above a certain velocity (i.e. $v\geq 1.5 \ v_F$), not only binary but also collective excitations (plasmons) are possible \cite{mon17}. Moreover, at high energies, both the FEG and the bound electrons contribute to the stopping power. The method used in this work combines a description for the interaction with the valence (or conduction) electrons as a FEG, and a different one for the interaction with the bound electrons.

We used the SPCC model \cite{mon17} to describe the stopping power of low velocity charged particles in the FEG. The model relies on a non-perturbative binary collisional approximation. The SPCC \cite{mon17} is based on a screened central potential with a cusp condition of the electronic density close to the projectile. This model proved to give a good description of the induced electron density even for negative projectiles \cite{mon17} and reproduces the low-velocity proton-antiproton differences in the stopping power (Barkas effect). The SPCC formalism only depends on the Wigner-Seitz radio, $r_S$, which is a measure of the electronic density of the FEG. For metals of well-known $r_S$ values, the SPCC describes correctly the low energy experimental stopping data \cite{mon17}, agreeing with the DFT results by Echenique and coworkers \cite{eche81,nagy89} at $v=0$.

Hafnium ($Z=72$, [Xe] $4f^{14} \ 6s^2 \ 5d^2$) belongs to the first group of transition metals, with  4 electrons as FEG (theoretical $r_S=2.14$ a.u.) and $1s-4f$ electrons bound. We compared the computed $r_S$ with the experimental value obtained from the measured energy loss function by Lynch \textit{et al} \cite{lynch75}. The experimental plasmon energy of Hf is $\hbar\omega_P \approx 15.8$ eV, with a width at half maximum of $\delta \approx 4.4$ eV, and $r_S \approx 2.07$ a.u. \cite{lynch75}. The difference of less than $5\%$ between theoretical and experimental $r_S$ assess Hf as a canonical target\cite{mon17}.

Above certain impact velocity, the plasmon contribution is important (i.e. around and above the stopping maximum). An interesting value for our analysis is the minimum impact velocity to excite
plasmons, $v_P$. In the dielectric formalism, this value can be obtained as $v_P\ \approx \ v_F[1+(3 \pi\ v_F)^{-1/2} ]$ \cite{suppression}. To describe the energy loss considering collective and binary excitation, we resort to the ML dielectric formalism \cite{Mermin}, which is a linear response, perturbative approximation, so it depends on the square of the ion charge. In this formalism, the response of target electrons to the ion passage is described through the quantum dielectric function, depending on the characteristic $r_S$ and $\delta$ parameters of the FEG.

@@

For the stopping power due to bound electrons, the SLPA \cite{mon17,mon13} is employed. It is worth mentioning that the only inputs for the SLPA are the space-dependent densities of each shell in the ground state, and their binding energies. Collective processes and screening among electrons are included. Hafnium is a relativistic target, therefore the wave functions and binding energies must be obtained as solutions of the Dirac equation\cite{mendez2019}.

To asses the importance of the relativistic description of the bound electrons, Figure \ref{Binding_E} (a) shows non relativistic and relativistic binding energies $E_{nl\pm}$, where $\pm=j\pm1/2$.
We compare our results with experimental measurements on solid Hf \cite{williams1995} from $1s$ to $4f_{\pm}$. We notice that not only the most inner shells require relativistic calculations, but also the outer $5p$ and $4f$ shells. Figure \ref{Binding_E} (b) presents the relative error of non relativistic and relativistic energies with respect to the experimental data in logarithmic scale. This figure shows very clearly the disability of non-relativistic calculations to describe the experimental data, which surprisingly worsens from the inner to the outer shells. 
%It is clear in this figure the disability of non-relativistic binding energies to describe the experimental data, which surprisingly worsens from the inner to the outer-shells.
%We used here the theoretical binding energies and wave functions of Hf calculated in \cite{mendez2019}. 
%The present theoretical description of Hf considers a FEG of $r_S=2.14$ and 1s-4f electrons as bound.
